---
title: "Adventures in Covariance"
output: github_document
---

* Consider fitting a model to estimate the wait time at a variety of cafes. We might estimate that the average wait time varies by cafe, and can represent with a hierarchical model:

$$
\begin{gather}
\mu_i = \alpha_{\text{CAFE}[i]}
\end{gather}
$$

* The time of day may matter as well --- if all cafes experience a similar slow down in the afternoon ($A_i$), we can model using our previous methods:

$$
\begin{gather}
\mu_i = \alpha_{\text{CAFE}[i]} + \beta A_i
\end{gather}
$$

* If, however, different cafes had different changes in the afternoon, the slope parameter can also be modeled hierarchically:

$$
\begin{gather}
\mu_i = \alpha_{\text{CAFE}[i]} + \beta_{\text{CAFE}[i]} A_i
\end{gather}
$$

* This is also known as a *varying effects* model.
* Here's a fact that will allow us to squeeze more info out of the data: there is *covariance* in the intercept/slopes!
* For example, at a really popular cafe, the wait will be really long in the morning, but less so in the afternoon. At a bone-dry cafe, the difference between morning and afternoon will be less noticeable, because you can expect a short wait time even in the morning.
* We can use this covariance to pool information across parameter types --- intercepts and slopes.
* This chapter will explore the varying effects strategy & extend to more subtle model types, such as *Gaussian Processes*. 
* Ordinary varying effects only work with discrete, unordered categories like country or individual, but we can also use pooling with continuous categories like age or location. 
* We'll also toe-dip back into causal inference and introduce *instrumental variables*, which allow ways of inferring cause without closing backdoor paths in a DAG. 
* The material in this chapter is difficult!

## 14.1 Varying slopes by construction

* A joint multivariate Gaussian distribution is generally used to model the covariance of the intercepts and slopes.
* There's not inherent reason why the multivariate distribution of slopes/intercepts must be Gaussian, but there are practical and epidemiological reasons:
  * **practical**: there aren't many multivariate distributions that are easy to work with --- the only common ones are the MV Gaussian & MV Student-t.
  * **epistemological**: if all we want to say about the intercepts/slopes is their means, variances, and covariances, then the Gaussian is the maximum entropy distribution.
* Let's simulate some data from coffee shops to get a better idea.

### 14.1.1 Simulate the population

```{r}
a <- 3.5        # average morning wait time
b <- -1         # average difference in afternoon wait time
sigma_a <- 1    # std dev in intercepts
sigma_b <- 0.5  # std dev in slopes
rho <- -0.7     # correlation between intercepts and slopes
```

* To use these values to simulate a sample of cafes, we need a 2-d MV Gaussian which requires:
  1. a vector of 2 means
  2. a 2-by-2 matrix of variances and covariances
* The means are easy:

```{r}
Mu <- c(a, b)
```

* The matrix of variances is arranged like this:

$$
\begin{gather}
\begin{pmatrix}
\text{variance of intercepts} & \text{covariance of intercepts and slopes} \\
\text{covariance of intercepts and slopes} & \text{variance of slopes}
\end{pmatrix}
\end{gather}
$$

$$
\begin{gather}
\begin{pmatrix}
\sigma_{\alpha}^2 & \sigma_{\alpha} \sigma_{\beta} \rho \\
\sigma_{\alpha} \sigma_{\beta} \rho & \sigma_{\beta}^2
\end{pmatrix}
\end{gather}
$$

* We can build this covariance matrix in a few ways. The first is to use `matrix()`, which takes a vector and starts filling in a matrix by column:

```{r}
# here's how the matrix is filled
matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2)

# construct the matrix directly
cov_ab <- sigma_a * sigma_b * rho
Sigma <- matrix(c(sigma_a^2, cov_ab, cov_ab, sigma_b^2), ncol = 2)
Sigma
```

* Alternatively, we can treat the standard deviations and covariances separately then use matrix multiplication to combine:

```{r}
sigmas <- c(sigma_a, sigma_b) # std devs
Rho <- matrix(c(1, rho, rho, 1), nrow = 2) # correlation matrix
Sigma <- diag(sigmas) %*% Rho %*% diag(sigmas)
Sigma
```

* To simulate the properties of each cafe, we just sample randomly from the MV Gaussian:

```{r}
library(rethinking)

# simulate intercepts/slopes
N_cafes <- 20
set.seed(5)
vary_effects <- MASS::mvrnorm(N_cafes, Mu, Sigma)
vary_effects

# assign to vectors
a_cafe <- vary_effects[,1]
b_cafe <- vary_effects[,2]

# plot!
plot(a_cafe, b_cafe,
     col = rangi2,
     xlab = "intercepts (`a_cafe`)",
     ylab = "slopes (`b_cafe`)")

# overlay population distribution
for (i in c(0.1, 0.3, 0.5, 0.8, 0.99))
  lines(ellipse::ellipse(Sigma, centre = Mu, level = i), 
        col = col.alpha("black", 0.2))
```












