---
title: "The Many Variables & The Spurious Waffles"
output: github_document
math: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 500,
  fig.width = 9,
  fig.height = 6
)
```

* States with the highest number of Waffle Houses also have the highest divorce rate.
* This is a spurious relationship!
* *Multiple regression* helps deflect spurious relationships (more on that later...):
  * "Control" for confounds
  * Multiple and complex causation
  * Interactions

## 5.1 Spurious association

* The rate that adults marry is a great predictor of divorce rate, but does marriage *cause* divorce?

```{r}
library(rethinking)
data("WaffleDivorce")
d <- WaffleDivorce

# standardize variables
d$D <- standardize(d$Divorce)
d$M <- standardize(d$Marriage)
d$A <- standardize(d$MedianAgeMarriage)
```

$$
D_i \sim Normal(\mu_i, \sigma) \\
\mu_i = \alpha + \beta_A A_i \\
\alpha \sim Normal(0, 0.2) \\
\beta_A \sim Normal(0, 0.5) \\
\sigma \sim Exponential(1)
$$

```{r}
# model
m5.1 <-
  quap(
    alist(D ~ dnorm(mu, sigma),
          mu <- a + bA * A,
          a ~ dnorm(0, 0.2),
          bA ~ dnorm(0, 0.5),
          sigma ~ dexp(1)),
    data = d
  )

# simulate from priors:
set.seed(10)
prior <- extract.prior(m5.1)
mu <- link(m5.1, post = prior, data = list(A = c(-2, 2)))

# plot!
plot(NULL, 
     xlim = c(-2,2), 
     ylim = c(-2,2),
     xlab = "Median age marriage (std)",
     ylab = "Divorce rate (std)",
     main = "Prior simulation")

for (i in 1:50) lines(c(-2,2), mu[i,], col = col.alpha("black", 0.4))

# plotting function for posterior:


# compute percentile interval of mean
A_seq <- seq(from = -3, to = 3.2, length.out = 30)
mu <- link(m5.1, data = list(A = A_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# plot it all!
plot(D ~ A,
     data = d,
     col = rangi2,
     xlab = "Median age marriage (std)",
     ylab = "Divorce rate (std)",
     main = "Posterior")

lines(A_seq,
      mu.mean,
      lwd = 2)

shade(mu.PI, A_seq)

# model divorce rate as a fn of marriage rate
m5.2 <-
  quap(
    alist(D ~ dnorm(mu, sigma),
          mu <- a + bM * M,
          a ~ dnorm(0, 0.2),
          bM ~ dnorm(0, 0.5),
          sigma ~ dexp(1)),
    data = d
  )

# plot posterior for m5.2
M_seq <- seq(from = -2, to = 3, length.out = 30)
mu <- link(m5.2, data = list(M = M_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

plot(D ~ M,
     data = d,
     col = rangi2,
     xlab = "Marriage rate (std)",
     ylab = "Divorce rate (std)",
     main = "Posterior")

lines(M_seq,
      mu.mean,
      lwd = 2)

shade(mu.PI, M_seq)
```

* There is a weakly positive relationship between marriage and divorce rates, and a negative relationship between median age at marriage and divorce rate.
* Comparing these single-variable models separately isn't great --- they could both provide value, be redundant, or one can eliminate the value of the other.
* Goal: think *causally*, then fit a bigger regression.

### 5.1.1 Think before you regress

* Three variables at play: D, M, and A. 
* A *Directed Acyclic Graph* (DAG) can help us think about the relationship between the variables.
* A possible DAG for the divorce rate example could be:

```{r}
library(dagitty)
dag5.1 <- dagitty("dag{A -> D; A -> M; M -> D}")
coordinates(dag5.1) <- 
  list(x = c(A = 0, D = 1, M = 2),
       y = c(A = 0, D = 1, M = 0))

drawdag(dag5.1)
```

* In this diagram, age influences divorce rates in two ways: directly (A -> D) and indirectly through marriage rates (A -> M -> D)
* To infer the strength of these different arrows, we'd need more than one model. `m5.1` only shows the *total* influence of age at marriage on the divorce rate (direct and indirect). It could possible that there is no direct effect and it is associated with D entirely through the indirect path. This is known as *mediation*. 
* Another alternative is that there is no relationship between M & D (this is still consistent with `m5.2`, because M in this DAG picks up information from A):

```{r}
dag5.2 <- dagitty("dag{A -> M; A -> D}")
coordinates(dag5.2) <- coordinates(dag5.1)
drawdag(dag5.2)
```

* So, which is it?





